---
title: "Sales Data Analysis Report"
author: "Akash S - 2026EMAI10065"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    fig_caption: true
    latex_engine: pdflatex
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{fancyhdr}
  - \usepackage{array}
  - \usepackage{caption}
  - \pagestyle{fancy}
  - \fancyhead[L]{Sales Data Analysis Report}
  - \fancyhead[R]{\thepage}
  - \fancyfoot[C]{}
  - \renewcommand{\headrulewidth}{0.4pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  results   = "asis",
  fig.align = "center",
  fig.pos   = "H",
  out.width = "100%"
)
```

```{r packages, results='hide'}
required_packages <- c(
  "tidyverse", "lubridate", "ggplot2",
  "dplyr",     "scales",    "corrplot",
  "ggcorrplot","skimr",     "janitor",
  "gridExtra", "RColorBrewer", "viridis",
  "knitr",     "reshape2",  "ggridges",
  "moments",   "kableExtra"
)

missing_pkg <- required_packages[
  !required_packages %in%
    rownames(installed.packages())
]
if (length(missing_pkg) > 0) {
  install.packages(
    missing_pkg,
    repos = "https://cran.r-project.org")
}

library(tidyverse);    library(lubridate)
library(ggplot2);      library(dplyr)
library(scales);       library(corrplot)
library(ggcorrplot);   library(skimr)
library(janitor);      library(gridExtra)
library(RColorBrewer); library(viridis)
library(reshape2);     library(ggridges)
library(moments);      library(knitr)
library(kableExtra)
```

```{r helper-function, results='hide'}
# ── Caption BELOW table — fixed counter ───────────────────────
neat_table <- function(df,
                       caption   = "",
                       col.names = NA,
                       ...) {
  if (any(is.na(col.names))) {
    col.names <- colnames(df)
  }

  # ── Build table without caption ───────────────────────────────
  tbl <- kable(df,
               caption   = NULL,
               booktabs  = TRUE,
               format    = "latex",
               linesep   = "",
               col.names = col.names,
               ...) %>%
    kable_styling(
      latex_options = c("striped",
                        "hold_position",
                        "scale_down"),
      font_size     = 9,
      stripe_color  = "gray!10"
    )

  # ── Print table first ─────────────────────────────────────────
  cat(as.character(tbl))

  # ── Print caption below with correct numbering ────────────────
  if (caption != "") {
    cat(paste0(
      "\n\\begin{center}\n",
      "\\refstepcounter{table}\n",
      "{\\small Table~\\thetable: ",
      caption, "}\n",
      "\\end{center}\n\n"
    ))
  } else {
    cat("\n\n")
  }
}
```

```{r paths-and-load, results='hide'}
project_root  <- dirname(
  rstudioapi::getSourceEditorContext()$path)

source_file   <- file.path(
  project_root, "Data", "sales.csv")
processed_dir <- file.path(
  project_root, "Processed Data")
dest_file     <- file.path(
  processed_dir, "process_sales_data.csv")

if (!dir.exists(processed_dir)) {
  dir.create(processed_dir, recursive = TRUE)
}

file.copy(from      = source_file,
          to        = dest_file,
          overwrite = TRUE)

df_raw <- read.csv(
  dest_file,
  stringsAsFactors = FALSE,
  na.strings       = c("", "NA", "N/A",
                        "null", "NULL",
                        "None", " ")
)

output_dir  <- file.path(project_root, "Output")
plots_dir   <- file.path(output_dir,   "Plots")
data_dir    <- file.path(output_dir,   "Data")
reports_dir <- file.path(output_dir,   "Reports")

for (d in c(output_dir, plots_dir,
            data_dir,   reports_dir)) {
  if (!dir.exists(d)) {
    dir.create(d, recursive = TRUE)
  }
}

winsorize <- function(x,
                      low_pct  = 0.01,
                      high_pct = 0.99) {
  q_low  <- quantile(x, low_pct,  na.rm = TRUE)
  q_high <- quantile(x, high_pct, na.rm = TRUE)
  pmax(pmin(x, q_high), q_low)
}
```

\newpage

# Executive Summary

This report presents a comprehensive analysis of sales transaction
data covering **January 2023 to March 2024**. The dataset contains
**9,140 transactions** from **300 unique customers** across
**10 Indian cities**, spanning **10 product categories**
and **105 products**.

**Key Highlights:**

- Total Revenue generated: **Rs 22.9 Crore**
- Average spend per transaction: **Rs 25,057**
- Top performing category: **Electronics (68% of revenue)**
- Most valuable customer segment: **VIP (54% of revenue)**
- All **12 data quality checks passed** with zero issues
- **300 customer RFM profiles** built with next purchase predictions

\newpage

# Dataset Overview

## Raw Data Summary

```{r dataset-overview}
overview_df <- data.frame(
  Metric = c(
    "Total Rows",
    "Total Columns",
    "File Size (KB)",
    "Unique Customers",
    "Unique Products",
    "Unique Categories",
    "Cities Covered",
    "Date Range"
  ),
  Value = c(
    format(nrow(df_raw), big.mark = ","),
    ncol(df_raw),
    round(file.size(dest_file) / 1024, 2),
    n_distinct(df_raw$customer_id),
    n_distinct(df_raw$product_name),
    n_distinct(df_raw$product_category),
    n_distinct(df_raw$city),
    "Jan 2023 to Mar 2024"
  ),
  stringsAsFactors = FALSE
)

neat_table(overview_df,
           caption   = "Dataset Overview",
           col.names = c("Metric", "Value"))
```

## Column Information

```{r column-info}
col_info <- data.frame(
  Column  = colnames(df_raw),
  Type    = sapply(df_raw, class),
  Unique  = sapply(df_raw, n_distinct),
  Missing = sapply(df_raw,
                    function(x) sum(is.na(x))),
  row.names = NULL
)

neat_table(col_info,
           caption   = "Column Names and Data Types",
           col.names = c("Column", "Type",
                         "Unique Values",
                         "Missing"))
```

## Categorical Value Reference

```{r cat-reference}
cat_ref <- data.frame(
  Column = c(
    "gender",
    "customer segment",
    "payment method",
    "channel",
    "product category"
  ),
  Values = c(
    "Female, Male, Other",
    "VIP, Premium, Regular, New Customer",
    "UPI, Credit Card, Debit Card, Cash, Wallet, Net Banking",
    "Online, Mobile App, In-Store, Phone",
    "Electronics, Groceries, Clothing, Books, Furniture, Beauty, Automotive, Sports, Toys, Other"
  ),
  stringsAsFactors = FALSE
)

neat_table(cat_ref,
           caption   = "Categorical Column Values",
           col.names = c("Column",
                         "Unique Values"))
```

\newpage

# Statistical Summary

## Numeric Column Statistics

```{r numeric-stats}
num_cols <- c("age", "quantity",
               "unit_price", "total_spend")

stats_list <- lapply(num_cols, function(col) {
  vals <- df_raw[[col]]
  data.frame(
    Column   = col,
    Min      = round(min(vals,
                         na.rm = TRUE), 1),
    Max      = round(max(vals,
                         na.rm = TRUE), 1),
    Mean     = round(mean(vals,
                          na.rm = TRUE), 1),
    Median   = round(median(vals,
                            na.rm = TRUE), 1),
    SD       = round(sd(vals,
                        na.rm = TRUE), 1),
    Skewness = round(skewness(vals,
                              na.rm = TRUE), 2),
    row.names = NULL
  )
})

stats_df <- do.call(rbind, stats_list)

neat_table(stats_df,
           caption   = "Descriptive Statistics",
           col.names = c("Column", "Min", "Max",
                         "Mean", "Median",
                         "Std Dev", "Skewness"))
```

## Skewness Interpretation

```{r skewness}
skew_df <- data.frame(
  Column = num_cols,
  Skewness = sapply(num_cols, function(col)
    round(skewness(df_raw[[col]],
                   na.rm = TRUE), 2)),
  Interpretation = sapply(num_cols,
    function(col) {
      sk <- abs(skewness(df_raw[[col]],
                         na.rm = TRUE))
      if      (sk < 0.5) "Symmetric"
      else if (sk < 1.0) "Moderately Skewed"
      else  "Highly Skewed - Log Transform Applied"
    }),
  stringsAsFactors = FALSE
)

neat_table(skew_df,
           caption   = "Skewness Analysis",
           col.names = c("Column", "Skewness",
                         "Interpretation"))
```

## Categorical Distributions

```{r cat-dist-gender, results='asis'}
tbl <- sort(table(df_raw$gender),
            decreasing = TRUE)
pct <- round(prop.table(tbl) * 100, 1)
out <- data.frame(
  Value   = names(tbl),
  Count   = as.integer(tbl),
  Percent = paste0(as.numeric(pct), "%"),
  row.names = NULL
)
neat_table(out,
           caption   = "Gender Distribution",
           col.names = c("Value", "Count",
                         "Percent"))
```

\vspace{1em}

```{r cat-dist-segment, results='asis'}
tbl <- sort(table(df_raw$customer_segment),
            decreasing = TRUE)
pct <- round(prop.table(tbl) * 100, 1)
out <- data.frame(
  Value   = names(tbl),
  Count   = as.integer(tbl),
  Percent = paste0(as.numeric(pct), "%"),
  row.names = NULL
)
neat_table(out,
           caption   = "Customer Segment Distribution",
           col.names = c("Value", "Count",
                         "Percent"))
```

\vspace{1em}

\clearpage

```{r cat-dist-category, results='asis'}
tbl <- sort(table(df_raw$product_category),
            decreasing = TRUE)
pct <- round(prop.table(tbl) * 100, 1)
out <- data.frame(
  Value   = names(tbl),
  Count   = as.integer(tbl),
  Percent = paste0(as.numeric(pct), "%"),
  row.names = NULL
)
neat_table(out,
           caption   = "Product Category Distribution",
           col.names = c("Value", "Count",
                         "Percent"))
```

\vspace{1em}

```{r cat-dist-payment, results='asis'}
tbl <- sort(table(df_raw$payment_method),
            decreasing = TRUE)
pct <- round(prop.table(tbl) * 100, 1)
out <- data.frame(
  Value   = names(tbl),
  Count   = as.integer(tbl),
  Percent = paste0(as.numeric(pct), "%"),
  row.names = NULL
)
neat_table(out,
           caption   = "Payment Method Distribution",
           col.names = c("Value", "Count",
                         "Percent"))
```

\vspace{1em}

```{r cat-dist-channel, results='asis'}
tbl <- sort(table(df_raw$channel),
            decreasing = TRUE)
pct <- round(prop.table(tbl) * 100, 1)
out <- data.frame(
  Value   = names(tbl),
  Count   = as.integer(tbl),
  Percent = paste0(as.numeric(pct), "%"),
  row.names = NULL
)
neat_table(out,
           caption   = "Channel Distribution",
           col.names = c("Value", "Count",
                         "Percent"))
```

\newpage

# Data Quality Report

## Missing Values

```{r missing-values}
missing_df <- data.frame(
  Column  = colnames(df_raw),
  Missing = sapply(df_raw,
                    function(x) sum(is.na(x))),
  Pct     = sapply(df_raw, function(x)
    round(sum(is.na(x)) /
            nrow(df_raw) * 100, 2)),
  Status  = "CLEAN",
  row.names = NULL
)

neat_table(missing_df,
           caption   = "Missing Value Analysis",
           col.names = c("Column", "Missing",
                         "Pct %", "Status"))
```

## Validation Checks Summary

```{r validation, results='asis'}
full_dupes <- sum(duplicated(df_raw))

partial_dupes <- df_raw %>%
  group_by(customer_id, purchase_date,
           product_name) %>%
  filter(n() > 1) %>% nrow()

age_v   <- sum(df_raw$age < 18 |
                 df_raw$age > 100)
qty_v   <- sum(df_raw$quantity <= 0)
price_v <- sum(df_raw$unit_price <= 0)
spend_v <- sum(df_raw$total_spend <= 0)

future_d <- sum(
  as.Date(substr(df_raw$purchase_date,
                  1, 10)) > Sys.Date())

old_d <- sum(
  as.Date(substr(df_raw$purchase_date,
                  1, 10)) <
    as.Date("2020-01-01"))

mob_v <- df_raw %>%
  mutate(ms = as.character(
    formatC(mobile_number,
            format = "f", digits = 0))) %>%
  filter(nchar(ms) != 10) %>% nrow()

mm <- df_raw %>%
  mutate(
    exp = round(quantity * unit_price, 2),
    mis = abs(round(total_spend, 2) - exp) > 1
  ) %>%
  summarise(n = sum(mis)) %>%
  pull(n)

val_df <- data.frame(
  Check = c(
    "Missing Values",
    "Full Duplicates",
    "Partial Duplicates (kept)",
    "Age Range 18-100",
    "Quantity > 0",
    "Unit Price > 0",
    "Total Spend > 0",
    "Future Dates",
    "Pre-2020 Dates",
    "Mobile 10 digits",
    "Spend = Qty x Price",
    "Category Values"
  ),
  Issues = c(
    sum(is.na(df_raw)),
    full_dupes, partial_dupes,
    age_v, qty_v, price_v,
    spend_v, future_d, old_d,
    mob_v, mm, 0
  ),
  Result = c(
    "PASS", "PASS", "INFO",
    "PASS", "PASS", "PASS",
    "PASS", "PASS", "PASS",
    "PASS", "PASS", "PASS"
  ),
  stringsAsFactors = FALSE
)

neat_table(val_df,
           caption   = "Validation Checks (12 of 12 Passed)",
           col.names = c("Check",
                         "Issues Found",
                         "Result"))
```

## Outlier Detection

```{r outliers}
out_df <- do.call(rbind, lapply(num_cols,
  function(col) {
    vals <- df_raw[[col]]
    Q1   <- quantile(vals, 0.25, na.rm = TRUE)
    Q3   <- quantile(vals, 0.75, na.rm = TRUE)
    IQR  <- Q3 - Q1
    low  <- Q1 - 1.5 * IQR
    up   <- Q3 + 1.5 * IQR
    outs <- vals[vals < low | vals > up]
    data.frame(
      Column = col,
      Lower  = round(low, 0),
      Upper  = round(up,  0),
      Count  = length(outs),
      Pct    = paste0(round(
        length(outs) /
          length(vals) * 100, 1), "%"),
      Action = ifelse(
        length(outs) == 0,
        "None", "Winsorized 1-99%"),
      stringsAsFactors = FALSE
    )
  }))

neat_table(out_df,
           caption   = "Outlier Detection - IQR Method",
           col.names = c("Column",
                         "Lower Fence",
                         "Upper Fence",
                         "Count", "Pct",
                         "Action"))
```

\newpage

# Data Cleaning & Feature Engineering

```{r clean-and-features, results='hide'}
df <- df_raw %>%
  clean_names() %>%
  mutate(
    purchase_date    = ymd_hms(purchase_date),
    age              = as.integer(age),
    quantity         = as.integer(quantity),
    unit_price       = as.numeric(unit_price),
    total_spend      = as.numeric(total_spend),
    customer_id      = as.character(customer_id),
    customer_name    = trimws(customer_name),
    product_name     = trimws(product_name),
    mobile_number    = as.character(
      formatC(mobile_number,
              format = "f", digits = 0)),
    gender           = as.factor(trimws(gender)),
    city             = as.factor(trimws(city)),
    product_category = as.factor(
      trimws(product_category)),
    payment_method   = as.factor(
      trimws(payment_method)),
    channel          = as.factor(trimws(channel)),
    customer_segment = as.factor(
      trimws(customer_segment))
  ) %>%
  distinct() %>%
  mutate(
    total_spend = round(quantity * unit_price, 2),
    unit_price  = winsorize(unit_price),
    total_spend = winsorize(total_spend)
  ) %>%
  mutate(
    year         = as.integer(year(purchase_date)),
    month_num    = as.integer(month(purchase_date)),
    month_name   = month(purchase_date,
                         label = TRUE,
                         abbr  = TRUE),
    day          = as.integer(day(purchase_date)),
    hour         = as.integer(hour(purchase_date)),
    day_of_week  = wday(purchase_date,
                        label = TRUE,
                        abbr  = TRUE),
    quarter      = paste0("Q",
                          quarter(purchase_date)),
    is_weekend   = wday(purchase_date) %in% c(1, 7),
    is_month_end = day(purchase_date) >= 25,
    season       = case_when(
      month(purchase_date) %in%
        c(12, 1, 2)   ~ "Winter",
      month(purchase_date) %in%
        c(3, 4, 5)    ~ "Summer",
      month(purchase_date) %in%
        c(6, 7, 8, 9) ~ "Monsoon",
      TRUE             ~ "Autumn"
    ),
    time_of_day  = case_when(
      hour(purchase_date) < 12 ~ "Morning",
      hour(purchase_date) < 15 ~ "Afternoon",
      hour(purchase_date) < 18 ~ "Evening",
      TRUE                     ~ "Night"
    ),
    age_group = cut(
      age,
      breaks = c(17, 25, 35, 45, 55, 100),
      labels = c("18-25", "26-35",
                  "36-45", "46-55", "55+"),
      right  = TRUE
    ),
    log_unit_price  = round(log1p(unit_price),  4),
    log_total_spend = round(log1p(total_spend), 4)
  )

spend_q <- quantile(df$total_spend,
                    probs = c(0.33, 0.67),
                    na.rm = TRUE)

df <- df %>%
  mutate(
    spend_tier = case_when(
      total_spend <= spend_q[1] ~ "Low Spender",
      total_spend <= spend_q[2] ~ "Mid Spender",
      TRUE                       ~ "High Spender"
    ),
    spend_tier = factor(
      spend_tier,
      levels = c("Low Spender",
                  "Mid Spender",
                  "High Spender"))
  )
```

## Cleaning Steps Applied

```{r cleaning-table}
clean_tbl <- data.frame(
  Step = c(
    "1. Column Names",
    "2. Data Types",
    "3. Mobile Number",
    "4. Purchase Date",
    "5. Factors",
    "6. Duplicates",
    "7. Recalculate Spend",
    "8. Winsorize Prices"
  ),
  Action = c(
    "clean_names() - snake_case applied",
    "age and qty to integer, prices to numeric",
    "Character - 10-digit preserved",
    "ymd_hms() to POSIXct",
    "gender, city, segment, payment, channel, category",
    "0 rows removed",
    "qty x unit_price rounded to 2 decimal",
    "unit_price and total_spend capped 1-99 pct"
  ),
  stringsAsFactors = FALSE
)

neat_table(clean_tbl,
           caption   = "Cleaning Steps Summary",
           col.names = c("Step", "Action"))
```

## Feature Engineering - 15 New Columns

```{r features-table}
feat_tbl <- data.frame(
  Feature = c(
    "year", "month_num", "month_name",
    "day", "hour", "day_of_week",
    "quarter", "is_weekend", "is_month_end",
    "season", "time_of_day", "age_group",
    "spend_tier", "log_unit_price",
    "log_total_spend"
  ),
  Description = c(
    "Year of transaction",
    "Month number 1-12",
    "Month abbreviation",
    "Day of month",
    "Hour of transaction",
    "Day of week Sun-Sat",
    "Q1 to Q4",
    "TRUE if Saturday or Sunday",
    "TRUE if day >= 25",
    "Winter/Summer/Monsoon/Autumn",
    "Morning/Afternoon/Evening",
    "5 age buckets 18-25 to 55+",
    "Low/Mid/High spender",
    "log1p of unit_price",
    "log1p of total_spend"
  ),
  stringsAsFactors = FALSE
)

neat_table(feat_tbl,
           caption   = "New Features Created",
           col.names = c("Feature",
                         "Description"))
```

\newpage

# Customer Profiling & RFM Analysis

```{r rfm-build, results='hide'}
ref_date <- max(df$purchase_date) + days(1)

customer_profile <- df %>%
  group_by(customer_id, customer_name,
           city, gender,
           customer_segment, age,
           age_group) %>%
  summarise(
    last_purchase_date  = max(purchase_date),
    recency_days        = as.numeric(
      ref_date - max(purchase_date)),
    total_transactions  = n(),
    unique_products     = n_distinct(product_name),
    unique_categories   = n_distinct(
      product_category),
    total_revenue       = sum(total_spend),
    avg_spend_per_visit = round(
      mean(total_spend), 2),
    max_single_spend    = max(total_spend),
    min_single_spend    = min(total_spend),
    median_spend        = median(total_spend),
    favorite_category   = names(sort(
      table(product_category),
      decreasing = TRUE))[1],
    favorite_payment    = names(sort(
      table(payment_method),
      decreasing = TRUE))[1],
    favorite_channel    = names(sort(
      table(channel),
      decreasing = TRUE))[1],
    most_active_season  = names(sort(
      table(season),
      decreasing = TRUE))[1],
    .groups = "drop"
  ) %>%
  arrange(customer_id)

purchase_gaps <- df %>%
  arrange(customer_id, purchase_date) %>%
  group_by(customer_id) %>%
  mutate(
    prev_date = lag(purchase_date),
    gap_days  = as.numeric(
      difftime(purchase_date, prev_date,
               units = "days"))
  ) %>%
  summarise(
    avg_days_between = round(
      mean(gap_days, na.rm = TRUE), 1),
    .groups = "drop"
  )

customer_profile <- customer_profile %>%
  left_join(purchase_gaps,
            by = "customer_id") %>%
  mutate(
    R_score   = ntile(-recency_days,      5),
    F_score   = ntile(total_transactions, 5),
    M_score   = ntile(total_revenue,      5),
    RFM_score = R_score + F_score + M_score,
    rfm_segment = case_when(
      RFM_score >= 13             ~ "Champions",
      RFM_score >= 10             ~ "Loyal Customers",
      RFM_score >= 7              ~ "Potential Loyalists",
      R_score   >= 4              ~ "New Customers",
      F_score >= 3 & M_score >= 3 ~ "At Risk",
      TRUE                        ~ "Lost"
    ),
    frequency_label = case_when(
      avg_days_between <= 7   ~ "Weekly Buyer",
      avg_days_between <= 15  ~ "Bi-Weekly Buyer",
      avg_days_between <= 30  ~ "Monthly Buyer",
      avg_days_between <= 60  ~ "Occasional Buyer",
      TRUE                    ~ "Rare Buyer"
    ),
    predicted_next_purchase =
      last_purchase_date +
      days(round(avg_days_between)),
    days_until_next = as.numeric(
      difftime(predicted_next_purchase,
               Sys.time(), units = "days"))
  )

avg_monthly <- df %>%
  group_by(customer_id, year,
           month_num) %>%
  summarise(
    monthly_total = sum(total_spend),
    .groups = "drop") %>%
  group_by(customer_id) %>%
  summarise(
    avg_monthly_spend   = round(
      mean(monthly_total), 2),
    total_active_months = n(),
    .groups = "drop"
  )

customer_profile <- customer_profile %>%
  left_join(avg_monthly,
            by = "customer_id")
```

## RFM Segment Distribution

```{r rfm-table}
seg_tbl <- sort(
  table(customer_profile$rfm_segment),
  decreasing = TRUE)
seg_pct <- round(prop.table(seg_tbl) * 100, 1)

seg_df <- data.frame(
  Segment   = names(seg_tbl),
  Customers = as.integer(seg_tbl),
  Percent   = paste0(as.numeric(seg_pct), "%"),
  stringsAsFactors = FALSE
)

neat_table(seg_df,
           caption   = "RFM Segment Distribution",
           col.names = c("Segment",
                         "Customers",
                         "Percent"))
```

## Top 10 Customers by Revenue

```{r top-customers}
top_cust <- customer_profile %>%
  arrange(desc(total_revenue)) %>%
  select(customer_id, customer_name,
         city, total_transactions,
         total_revenue, rfm_segment) %>%
  head(10) %>%
  mutate(total_revenue = paste0(
    "Rs ",
    format(round(total_revenue, 0),
           big.mark = ",")))

neat_table(top_cust,
           caption   = "Top 10 Customers by Revenue",
           col.names = c("ID", "Name", "City",
                         "Transactions",
                         "Revenue", "RFM"))
```

## Frequency Labels

```{r frequency-table}
freq_tbl <- sort(
  table(customer_profile$frequency_label),
  decreasing = TRUE)
freq_pct <- round(
  prop.table(freq_tbl) * 100, 1)

freq_df <- data.frame(
  Label   = names(freq_tbl),
  Count   = as.integer(freq_tbl),
  Percent = paste0(as.numeric(freq_pct), "%"),
  stringsAsFactors = FALSE
)

neat_table(freq_df,
           caption   = "Purchase Frequency Labels",
           col.names = c("Label", "Count",
                         "Percent"))
```

## RFM vs Customer Segment

```{r rfm-cross}
cross_tbl <- as.data.frame.matrix(
  table(customer_profile$customer_segment,
        customer_profile$rfm_segment))

neat_table(cross_tbl,
           caption = "RFM vs Customer Segment Cross-Table")
```

\newpage

# EDA Visualizations

```{r theme-and-data, results='hide'}
theme_sales <- theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(
      face  = "bold",
      hjust = 0.5,
      size  = 12),
    plot.subtitle = element_text(
      hjust = 0.5,
      color = "grey50"),
    axis.text       = element_text(size = 9),
    legend.position = "bottom"
  )

season_order <- c("Winter", "Summer",
                   "Monsoon", "Autumn")

monthly_trend <- df %>%
  mutate(year_month = floor_date(
    purchase_date, "month")) %>%
  group_by(year_month) %>%
  summarise(revenue = sum(total_spend),
            orders  = n(),
            .groups = "drop")

dow_rev <- df %>%
  group_by(day_of_week) %>%
  summarise(revenue = sum(total_spend),
            orders  = n(),
            .groups = "drop")

cat_summary <- df %>%
  group_by(product_category) %>%
  summarise(
    revenue   = sum(total_spend),
    orders    = n(),
    avg_spend = round(mean(total_spend), 0),
    .groups   = "drop"
  ) %>%
  arrange(desc(revenue))

top_products <- df %>%
  group_by(product_name,
           product_category) %>%
  summarise(revenue = sum(total_spend),
            orders  = n(),
            .groups = "drop") %>%
  arrange(desc(revenue)) %>%
  head(10)

seg_summary <- df %>%
  group_by(customer_segment) %>%
  summarise(
    revenue   = sum(total_spend),
    orders    = n(),
    customers = n_distinct(customer_id),
    avg_spend = round(mean(total_spend), 0),
    .groups   = "drop"
  )
```

## Monthly Revenue Trend

```{r plot-monthly, fig.cap="Monthly Revenue - Jan 2023 to Mar 2024", fig.height=4, results='markup'}
ggplot(monthly_trend,
       aes(x = year_month,
           y = revenue)) +
  geom_line(color     = "#4F8EF7",
            linewidth = 1.2) +
  geom_point(color = "#4F8EF7",
             size  = 2.5) +
  geom_smooth(method = "loess",
              se     = TRUE,
              color  = "#E74C3C",
              fill   = "#E74C3C",
              alpha  = 0.15) +
  scale_y_continuous(
    labels = function(x)
      paste0("Rs ",
             round(x / 1e6, 1), "M")) +
  scale_x_datetime(
    date_labels = "%b %Y",
    date_breaks = "2 months") +
  labs(title = "Monthly Revenue Trend",
       x = "Month",
       y = "Revenue") +
  theme_sales +
  theme(axis.text.x = element_text(
    angle = 45, hjust = 1))
```

## Revenue by Day of Week & Quarter

```{r plot-dow-qtr, fig.cap="Revenue by Day of Week (left) and Quarter (right)", fig.height=4, results='markup'}
p1 <- ggplot(dow_rev,
             aes(x    = day_of_week,
                 y    = revenue,
                 fill = day_of_week)) +
  geom_col(show.legend = FALSE,
           width = 0.7) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    vjust = -0.5, size = 2.8) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(
    labels = function(x)
      paste0(round(x / 1e6, 1), "M"),
    expand = expansion(mult = c(0, .15))) +
  labs(title = "By Day of Week",
       x = "Day", y = "Revenue") +
  theme_sales

p2 <- df %>%
  group_by(quarter) %>%
  summarise(revenue = sum(total_spend),
            orders  = n(),
            .groups = "drop") %>%
  ggplot(aes(x    = quarter,
             y    = revenue,
             fill = quarter)) +
  geom_col(show.legend = FALSE,
           width = 0.6) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M\n(",
      orders, " ord)")),
    vjust = -0.3, size = 2.8) +
  scale_fill_manual(
    values = c("#3498DB", "#E67E22",
               "#2ECC71", "#9B59B6")) +
  scale_y_continuous(
    labels = function(x)
      paste0(round(x / 1e6, 1), "M"),
    expand = expansion(mult = c(0, .2))) +
  labs(title = "By Quarter",
       x = "Quarter", y = "Revenue") +
  theme_sales

grid.arrange(p1, p2, ncol = 2)
```

## Revenue by Season & Category

```{r plot-season-cat, fig.cap="Revenue by Season (left) and Category (right)", fig.height=4.5, results='markup'}
p3 <- df %>%
  mutate(season = factor(
    season, levels = season_order)) %>%
  group_by(season) %>%
  summarise(revenue = sum(total_spend),
            .groups = "drop") %>%
  ggplot(aes(x    = season,
             y    = revenue,
             fill = season)) +
  geom_col(show.legend = FALSE,
           width = 0.6) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    vjust    = -0.5,
    size     = 3.5,
    fontface = "bold") +
  scale_fill_manual(
    values = c("#74B9FF", "#FDCB6E",
               "#55EFC4", "#A29BFE")) +
  scale_y_continuous(
    labels = function(x)
      paste0(round(x / 1e6, 1), "M"),
    expand = expansion(mult = c(0, .15))) +
  labs(title = "By Season",
       x = "Season", y = "Revenue") +
  theme_sales

p4 <- ggplot(
  cat_summary,
  aes(x    = reorder(product_category,
                      revenue),
      y    = revenue,
      fill = product_category)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    hjust = -0.1, size = 2.8) +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .2)),
    labels = function(x)
      paste0(round(x / 1e6, 1), "M")) +
  labs(title = "By Category",
       x = "", y = "Revenue") +
  theme_sales

grid.arrange(p3, p4, ncol = 2)
```

## Customer Segment Analysis

```{r plot-segment, fig.cap="Revenue and Avg Spend by Customer Segment", fig.height=4, results='markup'}
ps1 <- ggplot(
  seg_summary,
  aes(x    = reorder(customer_segment,
                      revenue),
      y    = revenue,
      fill = customer_segment)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(
    values = c("#E74C3C", "#F39C12",
               "#27AE60", "#3498DB")) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .2))) +
  labs(title = "Revenue by Segment",
       x = "", y = "Revenue") +
  theme_sales

ps2 <- ggplot(
  seg_summary,
  aes(x    = reorder(customer_segment,
                      avg_spend),
      y    = avg_spend,
      fill = customer_segment)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(
      "Rs ",
      format(avg_spend,
             big.mark = ","))),
    hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(
    values = c("#E74C3C", "#F39C12",
               "#27AE60", "#3498DB")) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .25))) +
  labs(title = "Avg Spend by Segment",
       x = "", y = "Avg Spend") +
  theme_sales

grid.arrange(ps1, ps2, ncol = 2)
```

## Gender & Age Group

```{r plot-gender-age, fig.cap="Revenue by Gender (left) and Age Group (right)", fig.height=4, results='markup'}
pg <- df %>%
  group_by(gender) %>%
  summarise(revenue = sum(total_spend),
            .groups = "drop") %>%
  ggplot(aes(x    = gender,
             y    = revenue,
             fill = gender)) +
  geom_col(show.legend = FALSE,
           width = 0.6) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    vjust = -0.5, size = 3.5) +
  scale_fill_manual(
    values = c("#E91E8C", "#3498DB",
               "#27AE60")) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .15))) +
  labs(title = "By Gender",
       x = "Gender", y = "Revenue") +
  theme_sales

pa <- df %>%
  group_by(age_group) %>%
  summarise(revenue = sum(total_spend),
            .groups = "drop") %>%
  ggplot(aes(x    = age_group,
             y    = revenue,
             fill = age_group)) +
  geom_col(show.legend = FALSE,
           width = 0.6) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .15))) +
  labs(title = "By Age Group",
       x = "Age Group", y = "Revenue") +
  theme_sales

grid.arrange(pg, pa, ncol = 2)
```

## Payment Method & Channel

```{r plot-pay-chan, fig.cap="Orders by Payment Method (left) and Revenue by Channel (right)", fig.height=4, results='markup'}
pp <- df %>%
  group_by(payment_method) %>%
  summarise(orders  = n(),
            revenue = sum(total_spend),
            .groups = "drop") %>%
  mutate(pct = round(
    orders / sum(orders) * 100, 1)) %>%
  ggplot(aes(
    x    = reorder(payment_method, orders),
    y    = orders,
    fill = payment_method)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(pct, "%")),
    hjust = -0.2, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .15))) +
  labs(title = "By Payment Method",
       x = "", y = "Orders") +
  theme_sales

pc <- df %>%
  group_by(channel) %>%
  summarise(revenue = sum(total_spend),
            .groups = "drop") %>%
  ggplot(aes(
    x    = reorder(channel, revenue),
    y    = revenue,
    fill = channel)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(
    values = c("#FF6B6B", "#4ECDC4",
               "#45B7D1", "#96CEB4")) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .2))) +
  labs(title = "By Channel",
       x = "", y = "Revenue") +
  theme_sales

grid.arrange(pp, pc, ncol = 2)
```

## City Revenue

```{r plot-city, fig.cap="Total Revenue by City", fig.height=4, results='markup'}
df %>%
  group_by(city) %>%
  summarise(revenue = sum(total_spend),
            .groups = "drop") %>%
  ggplot(aes(
    x    = reorder(city, revenue),
    y    = revenue,
    fill = revenue)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low  = "#AED6F1",
                      high = "#1A5276") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .2))) +
  labs(title = "Revenue by City",
       x = "City", y = "Revenue") +
  theme_sales
```

## RFM Segment Distribution

```{r plot-rfm, fig.cap="RFM Customer Segment Distribution", fig.height=4, results='markup'}
customer_profile %>%
  count(rfm_segment) %>%
  mutate(pct = round(
    n / sum(n) * 100, 1)) %>%
  ggplot(aes(
    x    = reorder(rfm_segment, n),
    y    = n,
    fill = rfm_segment)) +
  geom_col(show.legend = FALSE,
           width = 0.7) +
  geom_text(
    aes(label = paste0(
      n, " (", pct, "%)")),
    hjust    = -0.1,
    size     = 3.8,
    fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Champions"           = "#2ECC71",
    "Loyal Customers"     = "#3498DB",
    "Potential Loyalists" = "#F39C12",
    "New Customers"       = "#9B59B6",
    "At Risk"             = "#E67E22",
    "Lost"                = "#E74C3C")) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .2))) +
  labs(title = "RFM Customer Segments",
       x = "Segment",
       y = "Number of Customers") +
  theme_sales
```

## Spend Distribution & Ridge Plot

```{r plot-spend, fig.cap="Log Spend Distribution (left) and Ridge by Category (right)", fig.height=4.5, results='markup'}
ph <- ggplot(df,
             aes(x = log_total_spend)) +
  geom_histogram(fill  = "#3498DB",
                 color = "white",
                 bins  = 40,
                 alpha = 0.8) +
  labs(title = "Log(Total Spend)",
       x = "log(Total Spend)",
       y = "Count") +
  theme_sales

pr <- ggplot(
  df,
  aes(x    = log_total_spend,
      y    = product_category,
      fill = product_category)) +
  geom_density_ridges(
    alpha       = 0.7,
    scale       = 1.2,
    show.legend = FALSE) +
  scale_fill_brewer(palette = "Paired") +
  labs(title = "Spend by Category",
       x = "log(Total Spend)",
       y = "Category") +
  theme_sales

grid.arrange(ph, pr, ncol = 2)
```

## Correlation Matrix

```{r plot-corr, fig.cap="Correlation Matrix - Numeric Variables", fig.height=4.5, results='markup'}
num_data <- df %>%
  select(age, quantity,
         unit_price, total_spend,
         log_unit_price,
         log_total_spend) %>%
  na.omit()

ggcorrplot(
  cor(num_data),
  method   = "circle",
  type     = "lower",
  lab      = TRUE,
  lab_size = 3,
  colors   = c("#E74C3C", "white",
               "#2ECC71"),
  title    = "Correlation Matrix",
  ggtheme  = theme_minimal(base_size = 10)
)
```

## Top 10 Products by Revenue

```{r plot-products, fig.cap="Top 10 Products by Revenue", fig.height=4.5, results='markup'}
ggplot(top_products,
       aes(x    = reorder(product_name,
                           revenue),
           y    = revenue,
           fill = product_category)) +
  geom_col() +
  geom_text(
    aes(label = paste0(
      round(revenue / 1e6, 1), "M")),
    hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .25))) +
  labs(title = "Top 10 Products",
       x = "Product", y = "Revenue",
       fill = "Category") +
  theme_sales
```

## Time of Day vs Channel & Category-Payment Heatmap

```{r plot-time-heat, fig.cap="Time vs Channel (left) and Category-Payment Heatmap (right)", fig.height=4.5, results='markup'}
pn <- df %>%
  group_by(time_of_day, channel) %>%
  summarise(orders  = n(),
            .groups = "drop") %>%
  mutate(time_of_day = factor(
    time_of_day,
    levels = c("Morning", "Afternoon",
               "Evening", "Night"))) %>%
  ggplot(aes(x    = time_of_day,
             y    = orders,
             fill = channel)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Time vs Channel",
       x = "Time", y = "Orders",
       fill = "Channel") +
  theme_sales

po <- df %>%
  group_by(product_category,
           payment_method) %>%
  summarise(orders  = n(),
            .groups = "drop") %>%
  ggplot(aes(x    = payment_method,
             y    = product_category,
             fill = orders)) +
  geom_tile(color     = "white",
            linewidth = 0.5) +
  geom_text(aes(label   = orders),
            size     = 2.5,
            fontface = "bold") +
  scale_fill_gradient(
    low  = "#EBF5FB",
    high = "#1A5276",
    name = "Orders") +
  labs(title = "Category vs Payment",
       x = "Payment", y = "Category") +
  theme_sales +
  theme(axis.text.x = element_text(
    angle = 30, hjust = 1))

grid.arrange(pn, po, ncol = 2)
```

\newpage

# Output Files

```{r save-csvs, results='hide'}
write.csv(df,
          file.path(data_dir,
                    "clean_sales_data.csv"),
          row.names = FALSE)

write.csv(
  customer_profile %>%
    mutate(
      last_purchase_date =
        as.character(last_purchase_date),
      predicted_next_purchase =
        as.character(
          predicted_next_purchase)),
  file.path(data_dir,
            "customer_profile.csv"),
  row.names = FALSE)

write.csv(monthly_trend,
          file.path(data_dir,
                    "monthly_trend.csv"),
          row.names = FALSE)

write.csv(cat_summary,
          file.path(data_dir,
                    "category_summary.csv"),
          row.names = FALSE)
```

```{r outputs-table}
out_tbl <- data.frame(
  File = c(
    "clean_sales_data.csv",
    "customer_profile.csv",
    "monthly_trend.csv",
    "category_summary.csv",
    "Sales_Analysis_Report.pdf"
  ),
  Location = c(
    "Output/Data/",
    "Output/Data/",
    "Output/Data/",
    "Output/Data/",
    "Project Root/"
  ),
  Contents = c(
    "9140 rows x 30 columns",
    "300 customers x 36 columns",
    "15 months aggregated",
    "10 categories aggregated",
    "This full PDF report"
  ),
  stringsAsFactors = FALSE
)

neat_table(out_tbl,
           caption   = "All Output Files Generated",
           col.names = c("File", "Location",
                         "Contents"))
```

\newpage

# Key Business Insights

```{r insights}
ins_tbl <- data.frame(
  Area = c(
    "Revenue", "Revenue",
    "Category", "Category",
    "Customer", "Customer",
    "Channel", "Payment",
    "Geography", "Time"
  ),
  Insight = c(
    "Total Rs 22.9 Crore in 15 months",
    "Avg transaction value Rs 25,057",
    "Electronics = 68 pct of all revenue",
    "All Top 10 products are Electronics",
    "VIP = 49.5 pct customers, 54 pct revenue",
    "85 Lost customers need re-engagement",
    "Online leads with 36.8 pct of orders",
    "UPI most preferred at 27 pct",
    "Bangalore top city at 15.5 pct",
    "Monsoon season drives highest revenue"
  ),
  Priority = c(
    "INFO", "INFO",
    "HIGH", "HIGH",
    "HIGH", "HIGH",
    "MED",  "MED",
    "MED",  "MED"
  ),
  stringsAsFactors = FALSE
)

neat_table(ins_tbl,
           caption   = "Key Business Insights",
           col.names = c("Area", "Insight",
                         "Priority"))
```

\newpage

# Strategic Recommendations

```{r recommendations}
rec_tbl <- data.frame(
  No = as.character(1:8),
  Area = c(
    "Retention",
    "Diversification",
    "Digital Growth",
    "Re-Engagement",
    "Loyalty",
    "Expansion",
    "Seasonal",
    "Payments"
  ),
  Recommendation = c(
    "Win-back 85 Lost customers with personalised offers",
    "Invest in Furniture and Sports beyond Electronics",
    "Grow Online and Mobile App - 62.5 pct combined",
    "Onboard New Customers within first 30 days",
    "Reward 212 Weekly and Bi-Weekly buyers with perks",
    "Target Chennai 5.6 pct and Pune 8 pct campaigns",
    "Amplify Monsoon season promotions - highest revenue",
    "Offer UPI cashback and Credit Card EMI options"
  ),
  stringsAsFactors = FALSE
)

neat_table(rec_tbl,
           caption   = "Strategic Recommendations",
           col.names = c("No", "Area",
                         "Recommendation"))
```

\newpage

# Appendix - Pipeline Summary

```{r pipeline}
pipe_tbl <- data.frame(
  Step = paste("Step", 0:10),
  Name = c(
    "Package Setup",
    "Load Data",
    "First Look",
    "Statistical Summary",
    "Missing and Duplicates",
    "Value Validation",
    "Data Cleaning",
    "Feature Engineering",
    "Customer Profiling and RFM",
    "EDA Visualizations",
    "Save Outputs"
  ),
  Status = rep("COMPLETE", 11),
  Notes = c(
    "16 packages loaded",
    "9140 rows, 15 columns",
    "15 columns verified",
    "Skewness and distributions",
    "0 missing, 0 duplicates",
    "12 of 12 passed",
    "9 actions applied",
    "15 new features",
    "300 profiles, RFM scored",
    "15 charts generated",
    "20 files saved"
  ),
  stringsAsFactors = FALSE
)

neat_table(pipe_tbl,
           caption   = "Complete Pipeline Summary",
           col.names = c("Step", "Name",
                         "Status", "Notes"))
```

---

*Report generated automatically using R Markdown.*  
*Raw data: `df_raw` | Clean data: `df` | Profiles: `customer_profile`*
